National Longitudinal Evidence for Growth in Subjective Wellbeing from Spiritual Beliefs

### 2020-11-21 Analysis.Rmd
added bayesian estimation

```{r global_options, echo=FALSE, include = FALSE}
knitr::opts_chunk$set(fig.path="Figs/", 
                      message=FALSE, 
                      warning=FALSE,
                      collapse =TRUE,
                      echo=TRUE, 
                      #results="hide", 
                      fig.width= 10,
                      fig.height=8,
                      tidy = "styler")
```
```{r load_libraries, echo=FALSE, include=FALSE}
library("styler")
library("tidyverse")
library("lme4")
library("ggplot2")
library("tidyr")
library("ggeffects")
library("Amelia")
library("table1")
library("patchwork")
library("easystats")
library("parameters")
library("prettycode")
library("table1")
library("gghighlight")
library("sjPlot")
library("texreg")
library("see")
#library("viridis")
#library("ggpubr")
#library("wesanderson")
library("ggsci")
library("papaja")
#library("here")
library("citr")

@@

Processing of data.
```{r data processing, cache = TRUE, dependson = "ldf.5"}
# filter only those who are measured 
dfben  <-ldf.5 %>%
  dplyr:: filter(Wave !=2009)%>% # exclude wave 1 (no spirit measures that wave)
  dplyr::filter(YearMeasured==1)%>%
  dplyr::group_by(Id) %>% filter(n() > 2)%>% # select those who have responded to at least 3 waves # Also works for 5 and 7
  dplyr::filter(n() !=0)%>%
  dplyr::ungroup(Id)%>%
  dplyr:: mutate(Years = as.numeric(years))%>%
  dplyr:: mutate(yearsC = scale(years, center=TRUE,scale=FALSE))%>%
  dplyr::mutate(Age.10yrs = (Age/10))%>%
  dplyr::mutate(Religion.Church.Log = log(Religion.Church+ 1))%>%
  dplyr::mutate(Pol.Orient.S = scale(Pol.Orient, center = TRUE, scale = TRUE),
         Age.10yrs.C = scale(Age.10yrs, center = TRUE, scale = FALSE),
         Male = factor(Gender),
         Employed = factor(Employed),
         Partner = factor(Partner),
         Edu.S = scale(as.numeric(Edu) ,scale=TRUE,center=TRUE),
         Education = as.numeric(Edu),
         Urban = factor(Urban),
         Believe.Spirit = as.factor(Believe.Spirit),
         Believe.God = as.factor(Believe.God),
         EthnicCategories = factor(EthnicCats),
         LIFESAT.S = scale(LIFESAT, center = TRUE, scale = TRUE),
         PWI.S = scale(PWI, center = TRUE, scale = TRUE),
         Deprivation.S = scale(NZdep, scale=TRUE, center=TRUE),
         Relid.C = scale(Relid, scale = TRUE,center=TRUE),
         Religious = as.factor(Religious),
         Religion.Church.Log.C = scale(Religion.Church.Log, center = TRUE, scale = FALSE))
```
```{r}
# nice labels
table1::label(dfben$Age.10yrs.C) <- "Age in Decades (C)"
table1::label(dfben$PWI) <- "Personal Wellbeing"
table1::label(dfben$LIFESAT) <- "Life Satisfaction"

@@

Table of belief types
```{r}
table(dfben$Beliefs)
range(dfben$years)
range(dfben$Years)
```

@@

Inference holds but VIFS years x beliefs when years are centered.

```{r model centering years}
library(lme4)
mod.1a <-lmer (PWI ~  1 +  yearsC * Beliefs +  Age.10yrs.C + Deprivation.S + Edu.S  + Employed + EthnicCats + Male  + Partner  + Pol.Orient.S + Urban + (1|Id), data= dfben, na.action = na.omit)
mod.2a <-lmer (LIFESAT ~ 1 + yearsC * Beliefs+ Age.10yrs.C + Deprivation.S + Edu.S  + Employed + EthnicCats + Male  + Partner  + Pol.Orient.S + Urban + (1|Id), data= dfben, na.action = na.omit)
summary(mod.1a)
summary(mod.2a)
# identical result
performance::check_collinearity(mod.1a)
performance::check_collinearity(mod.2a)

# address reviewer worry about deprivation and partner being correlated
m<-lmer (NZdep ~ Partner + (1|Id), data= dfben, na.action = na.omit)
tab_model(m)
model_parameters(m)
```

Model is robust without covariates

```{r robust to years and beliefs only}
library(lme4)
mod.1aa <-lmer (PWI ~  1 +  years * Beliefs + (1|Id), data= dfben, na.action = na.omit)
mod.2aa <-lmer (LIFESAT ~ 1 + years * Beliefs + (1|Id), data= dfben, na.action = na.omit)
)
# identical result
sjPlot::tab_model(mod.1aa,mod.2aa)
```

```{r cache = TRUE, dependson="mod.2", dependson="mod.1"}
tab_model(mod.2, mod.1) # Table
```

>>>>

```{r global_options, echo=FALSE, include = FALSE}
knitr::opts_chunk$set(fig.path="Figs/", 
                      message=FALSE, 
                      warning=FALSE,
                      collapse =TRUE,
                      echo=TRUE, 
                      #results="hide", 
                      fig.width= 10,
                      fig.height=8,
                      tidy = "styler")
```
```{r load_libraries, echo=FALSE, include=FALSE}
library("styler")
library("tidyverse")
library("lme4")
library("ggplot2")
library("tidyr")
library("ggeffects")
library("Amelia")
library("table1")
library("patchwork")
#library("easystats")
library("parameters")
library("prettycode")
library("table1")
library("gghighlight")
library("sjPlot")
library("texreg")
library("see")
#library("viridis")
#library("ggpubr")
#library("wesanderson")
library("ggsci")
library("papaja")
#library("here")
library(parameters) # great tables, plots
library("citr")
library("brms")
library("rstan")
rstan_options(auto_write=TRUE)
options(mc.cores=parallel::detectCores ())
#theme_set(theme_few())

@@

Processing of data.
```{r data processing, cache = TRUE, dependson = "ldf.5"}
# filter only those who are measured 
dfben  <-ldf.5 %>%
  dplyr:: filter(Wave !=2009)%>% # exclude wave 1 (no spirit measures that wave)
  dplyr::filter(YearMeasured==1)%>%
  dplyr::group_by(Id) %>% filter(n() > 2)%>% # select those who have responded to at least 3 waves # Also works for 5 and 7
  dplyr::filter(n() !=0)%>%
  dplyr::ungroup(Id)%>%
  dplyr:: mutate(Years = as.numeric(years))%>%
  dplyr:: mutate(yearsC = scale(years, center=TRUE,scale=FALSE))%>%
  dplyr::mutate(Age.10yrs = (Age/10))%>%
  dplyr::mutate(Religion.Church.Log = log(Religion.Church+ 1))%>%
  dplyr::mutate(Pol.Orient.S = scale(Pol.Orient, center = TRUE, scale = TRUE),
         Age.10yrs.C = scale(Age.10yrs, center = TRUE, scale = FALSE),
         Male = factor(Gender),
         Employed = factor(Employed),
         Partner = factor(Partner),
         Edu.S = scale(as.numeric(Edu) ,scale=TRUE,center=TRUE),
         Education = as.numeric(Edu),
         Urban = factor(Urban),
         Believe.Spirit = as.factor(Believe.Spirit),
         Believe.God = as.factor(Believe.God),
         EthnicCategories = factor(EthnicCats),
         LIFESAT.S = scale(LIFESAT, center = TRUE, scale = TRUE),
         PWI.S = scale(PWI, center = TRUE, scale = TRUE),
         Deprivation.S = scale(NZdep, scale=TRUE, center=TRUE),
         Relid.C = scale(Relid, scale = TRUE,center=TRUE),
         Religious = as.factor(Religious),
         Religion.Church.Log.C = scale(Religion.Church.Log, center = TRUE, scale = FALSE))

# nice labels
table1::label(dfben$Age.10yrs.C) <- "Age in Decades (C)"
table1::label(dfben$PWI) <- "Personal Wellbeing"
table1::label(dfben$LIFESAT) <- "Life Satisfaction"

@@

Table of belief types
```{r}
table(dfben$Beliefs)
# years
range(dfben$years)
```

@@

VIF high,  this is the years x factor interaction. See if it makes a difference by centering years.

```{r model centering years}
library(lme4)
mod.1a <-lmer (PWI ~  1 +  yearsC * Beliefs +  Age.10yrs.C + Deprivation.S + Edu.S  + Employed + EthnicCats + Male  + Partner  + Pol.Orient.S + Urban + (1|Id), data= dfben)
mod.2a <-lmer (LIFESAT ~ 1 + yearsC * Beliefs+ Age.10yrs.C + Deprivation.S + Edu.S  + Employed + EthnicCats + Male  + Partner  + Pol.Orient.S + Urban + (1|Id), data= dfben)
summary(mod.1a)
summary(mod.2a)
# identical result
performance::check_collinearity(mod.1a)
performance::check_collinearity(mod.2a)
```
Centering years makes no difference to inference.

Check reviewers interest in correlation of deprivation and having a partner.
```{r}
m<-lmer (NZdep ~ Partner + (1|Id), data= dfben, na.action = na.omit)
tab_model(m)
model_parameters(m)
```

Check whether model is robust without control covariates

```{r robust to years and beliefs only}
library(lme4)
mod.1aa <-lmer (PWI ~  1 +  yearsC * Beliefs + (1|Id), data= dfben)
mod.2aa <-lmer (LIFESAT ~ 1 + yearsC * Beliefs + (1|Id), data= dfben)
)
# identical result
sjPlot::tab_model(mod.1aa,mod.2aa)
```

Indeed, models are robust without control covariates
```{r cache = TRUE, dependson="mod.2", dependson="mod.1"}
tab_model(mod.2, mod.1) # Table
```

```{r}
# check model with bayesian estimator (preferred)
mod.1aaB <-brm(PWI ~  1 +  years * Beliefs + (1|Id), data= dfben, file ="mod.1aaB")
mod.2aaB <-brm (LIFESAT ~ 1 + years * Beliefs + (1|Id), data= dfben,
                file ="mod.2aaB")
```

note that the bayesian model recovers the time slope

```{r bayesian_pwi}
conditional_effects(mod.1aaB)[[3]]
```

```{r bayesian_lifesat}
conditional_effects(mod.2aaB)[[3]]

Try with a tensor spline

```{r}
# complete cases

dfbenTS<-dfben%>%
  select(PWI,LIFESAT,yearsC,Beliefs,Id)%>%
  filter(complete.cases(.))
```

```{r}
# check model with bayesian estimator (preferred)
mod.1aaB_ts <-brm(PWI ~  1 +  ts(years) + (1|Id), data= dfbenTS, file ="mod.1aaB_ts")

mod.2aaB_ts <-brm (LIFESAT ~ 1 + ts(Beliefs,years) + (1|Id), data= dfbenTS,
                file ="mod.2aaB_ts")
```

```{r}
str(dfben$Beliefs)
conditional_effects(mod.1aaB_ts, surface = TRUE)
conditional_smooths(mod.1aaB_ts)
```

```{r}
conditional_effects(mod.2aaB_ts, surface = TRUE)
```

```{r bayesian_pwi_plot }
pr.PWI_b <- ggpredict(model = mod.1aaB, terms = c("years [0:9]","Beliefs"), 
                    ci.lvl = 0.95,
                    type = "fe",
                    typical = "mean",
                   # condition = c(Euro =1),
                    back.transform = TRUE,
                    ppd = FALSE,
                    interval = "confidence")
p1<-plot(pr.PWI_b, facets = T) + 
 # scale_y_continuous(limits=c(4.75,5.25) )  + 
  gghighlight()  +  theme_blank()+ 
  ggtitle("Bayesian Model Predicted Values of Personal Wellbeing") 
p1
pr.PWI_b
```

```{r bayesian_pwi_plot }
pr.LS_b <- ggpredict(model = mod.2aaB, terms = c("years [0:9]","Beliefs"), 
                    ci.lvl = 0.95,
                    type = "fe",
                    typical = "mean",
                   # condition = c(Euro =1),
                    back.transform = TRUE,
                    ppd = FALSE,
                    interval = "confidence")
p1<-plot(pr.LS_b, facets = T) + 
 # scale_y_continuous(limits=c(4.75,5.25) )  + 
  gghighlight()  +  theme_blank()+ 
  ggtitle("Bayesian Model Predicted Values of Personal Wellbeing") 
p1
pr.LS_b
```

```{r bayesian_pwi_plot }
pr.PWI_b <- ggpredict(model = mod.21aaB, terms = c("years [0:9]","Beliefs"), 
                    ci.lvl = 0.95,
                    type = "fe",
                    typical = "mean",
                   # condition = c(Euro =1),
                    back.transform = TRUE,
                    ppd = FALSE,
                    interval = "confidence")
p1<-plot(pr.PWI_b, facets = T) + 
 # scale_y_continuous(limits=c(4.75,5.25) )  + 
  gghighlight()  +  theme_blank()+ 
  ggtitle("Vayesian Model Predicted Values of Personal Wellbeing") 
p1
pr.PWI_b
```

```{r bayesian_pwi_plot }
pr.LS_b <- ggpredict(model = mod.2aaB, terms = c("years [0:9]","Beliefs"), 
                    ci.lvl = 0.95,
                    type = "fe",
                    typical = "mean",
                   # condition = c(Euro =1),
                    back.transform = TRUE,
                    ppd = FALSE,
                    interval = "confidence")
p1<-plot(pr.LS_b, facets = T) + 
 # scale_y_continuous(limits=c(4.75,5.25) )  + 
  gghighlight()  +  theme_blank()+ 
  ggtitle("Bayesian Model Predicted Values of Personal Wellbeing") 
p1
pr.LS_b
```
